An Experimental Analysis of the Influence of the Precision of Parameters in an Artificial Neural Network

### 2025-05-28 paper/main.tex
first paper draft

% Packages
\usepackage[utf8]{inputenc}
\usepackage{amsmath, amssymb}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{geometry}
\usepackage{natbib}
\usepackage{setspace}
\usepackage{titlesec}

@@

	\bibliographystyle{plainnat}
	\bibliography{references}

\end{document}

>>>>

% Packages
\usepackage[utf8]{inputenc}
\usepackage{amsmath, amssymb}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{geometry}
\usepackage{setspace}
\usepackage{titlesec}
\usepackage{booktabs}
\usepackage{tabularx}
\usepackage{caption}
\usepackage{csquotes}
\usepackage{float}
\usepackage{derivative}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{hyperref}

\usepackage[style=nejm,
citestyle=numeric-comp,
sorting=none,
doi=false,
backend=bibtex]{biblatex}
\addbibresource{literature.bib}

@@

	\section{Background}

	A typical neuron in a neural network performs a weighted sum of its inputs, followed by an activation function:
	\[
	a = \omega^\top x + b,\quad \hat{y} = \sigma(a),
	\]
	where \( \omega \in \mathbb{R}^n \) is the weight vector, \( b \in \mathbb{R} \) is the bias, \( x \in \mathbb{R}^n \) is the input, and \( \sigma \) is a nonlinear activation function (e.g., sigmoid or ReLU).

	Training is conducted via backpropagation, where gradients of a loss function \( L(\hat{y}, y) \) with respect to the model parameters are computed. For example, using mean squared error:
	% \[
	% L(\hat{y}, y) = (\hat{y} - y)^2, \qquad \pdv{L}{\omega_i} = 2(\hat{y} - y) \cdot \odv{\sigma}{a} \cdot x_i.
	% \]

  	\begin{align*}
    	L(f(x_1, x_2)) &= (\sigma(\omega^T x + b) - y) ^2\\
                   	&= (\sigma(a) - y) ^ 2\\
                   	&= (\hat y - y) ^ 2
  	\end{align*}
  	\begin{alignat*}{3}
    	\pdv{L}{\omega_1} &= \pdv{L}{\sigma} &&\pdv{\sigma}{a}      &&\pdv{a}{\omega_1}\\
                      	&= 2(\hat y - y)~~ &&\odv{}{a}\sigma(a)~~ &&x_1
  	\end{alignat*}

	Memory and compute requirements during training are substantial, as gradients, activations, and parameters must be stored and updated iteratively. In contrast, inference requires only a forward pass, making it more amenable to optimization via precision reduction.

	\section{Memory Efficiency and Numerical Formats}

	To reduce memory usage and computational load, precision-aware techniques such as quantization and mixed-precision training are increasingly employed. Instead of using 32-bit or 64-bit floating-point formats, parameters can be stored in 16-bit or even 8-bit fixed-point representations.

	\begin{figure}[H]
    	\centering
    	\includegraphics[width=0.8\textwidth]{figures/float.png}
    	\caption{Illustration of standard floating-point format.}
	\end{figure}

	\begin{figure}[H]
    	\centering
    	\includegraphics[width=0.8\textwidth]{figures/binary-formats.png}
    	\caption{Comparison of binary formats used in neural network computations.}
	\end{figure}

	\section{Related Work}

	Prior research in this domain can be broadly classified into two categories:

	\subsection*{Quantization During Training}
	Several studies have explored training neural networks with reduced precision directly. Notable contributions include:
	\begin{itemize}
    	\item \textbf{Mixed Precision Training}, which leverages float16 arithmetic for speed while maintaining float32 master weights \cite{micikevicius2017mixed}.
    	\item \textbf{DoReFa-Net}, which quantizes weights, activations, and gradients to arbitrary bitwidths \cite{zhou2018dorefa}.
    	\item \textbf{Q-BERT}, a quantized BERT model that maintains accuracy using only 4-bit weights \cite{shen2020qbert}.
	\end{itemize}

	\subsection*{Post-Training Quantization}
	Other approaches focus on converting pretrained models into lower-precision equivalents:
	\begin{itemize}
    	\item \textbf{Integer-Only Quantization} using symmetric quantization scales \cite{jacob2018quantization}.
    	\item \textbf{GPTQ}, a method for quantizing large transformer models without retraining \cite{frantar2022gptq}.
    	\item \textbf{SmoothQuant}, which scales activations to better match weight ranges during quantization \cite{xiao2024smoothquant}.
	\end{itemize}

	\section{Experimental Design}

	We trained a simple fully connected neural network on the MNIST dataset, consisting of 60,000 training and 10,000 test grayscale images of handwritten digits (28×28 pixels). Model training was performed under varying parameter bitwidths, with cross-validation to ensure robustness.

	\begin{figure}[H]
    	\centering
    	\includegraphics[width=0.75\textwidth]{figures/mnist.png}
    	\caption{Sample images from the MNIST dataset.}
	\end{figure}

	\section{Results}

	\subsection*{Accuracy Across Bitwidths}

	\begin{figure}[H]
    	\centering
    	\includegraphics[width=0.75\textwidth]{figures/trainingAndInference.png}
    	\caption{Model accuracy during training and inference under different precision settings.}
	\end{figure}

	\subsection*{Training and Inference Time}

	\begin{figure}[H]
    	\centering
    	\includegraphics[width=0.75\textwidth]{figures/times.png}
    	\caption{Execution times for various precision configurations.}
	\end{figure}

	\subsection*{Quantization After Training}

	\begin{figure}[H]
    	\centering
    	\includegraphics[width=0.75\textwidth]{figures/inferenceOnly.png}
    	\caption{Inference accuracy using float64-trained weights followed by quantization.}
	\end{figure}

	\subsection*{Automatic Mixed Precision and Gradient Clipping}
	\begin{figure}[H]
    	\centering
    	\includegraphics[width=0.75\textwidth]{figures/gradclip_32.png}
    	\caption{Inference accuracy using float32-trained weights with AMP and gradient clipping.}
	\end{figure}

	\subsection*{Effect of Decimal Rounding}

	\begin{figure}[H]
    	\centering
    	\includegraphics[width=0.75\textwidth]{figures/118_accuracy.png}
    	\caption{Accuracy vs. number of decimal digits retained in weights.}
	\end{figure}

	\section{Discussion}

	Our results indicate that networks trained and evaluated with as low as 16-bit precision can achieve comparable performance to their high-precision counterparts. Moreover, rounding pretrained weights to just 3 decimal digits preserves full accuracy, opening the door to aggressive post-training quantization.

	This is particularly relevant for edge devices and real-time applications where memory and compute constraints are paramount.

	\section{Conclusion and Future Directions}

	This work confirms that for relatively simple tasks like MNIST classification, significant reductions in parameter precision can be achieved without compromising performance. Our future efforts will extend this analysis to more complex datasets (e.g., CIFAR-10, ImageNet) and deeper architectures, and explore advanced quantization-aware training techniques.

	\newpage
	\printbibliography
\end{document}
